Pool Model Paper

### 2023-12-06 pool_model.tex
Resource competition: rearrange, add Figures discussions

    \includegraphics[width=0.7\textwidth]{Figures/TPM_fig.pdf}

@@

Due to its life in rather unfavourable conditions bacteria developed different surviving mechanisms, one of which is adapting their metabolism and transitioning into dormant state.

@@

Further we show that the TPM can be readily extended to the systems with more than one bacterium and describe more complex relationships between species. 

@@

\label{Resource}

@@

Inserting Eq. \ref{Resource} into Eqs. \ref{eq:model_2sp_resource_comp} yields:

@@

which has the structure of a \ac{glv}.
In order to better see this, we set $L_A=L_B=0$: 

@@

Equations (\ref{eq:LV_simple1}-\ref{eq:LV_simple2}) are basically competitive Lotka-Volterra equations that are widely used in ecological studies and show the principle of competive excursion.
That principle claims that two species in competition for the same resource cannot coexist~\cite{wangersky_lotka-volterra_1978}. 
In previous studies it was shown that \ac{glv} equations with one bacterial state can be sucessfully applied to predict the microbial community dynamics, determine different types of the pairwise interspecies interactions~\cite{dedrick_when_2023, stein_ecological_2013, venturelli_deciphering_2018, hoffmann_power_2007}.
In food studies the \ac{glv} was applied by Mounier \etal~\cite{mounier_microbial_2008} to study interactions of six bacteria species  with yeast omissin in cheese.
Cauchie \etal~\cite{cauchie_modeling_2020} applied Lotka-Volterra model to study the three-species system of the bacteria majorly present in minced pork meat.

The pool model, in contrast to the classical Lotka-Volterra approach, allows for far more flexibility.

\begin{figure}[H]
    \begin{center}
    \includegraphics[width=0.67\textwidth]{Figures/pool_model_2pools_resource_competition.pdf}
    \caption{
        The competition of the two bacterial species due to the common limiting resource calculated using Eq. (\ref{eq:model_2sp_resource_comp}).
        The red curve describes the growth of the bacteria $A$ $N_A = \sum_{k} x_A^k = L_A+G_A$ with model parameters $\lambda_A=10^{-2}$, $\alpha_A=3.0$.
        The blue curve describes the growth of the bacteria $B$ $N_B = \sum_{k} x_B^k = L_B+G_B$ with model parameters $\lambda_A=5\cdot 10^{-2}$, $\alpha_A=2.5$.
        The purple line shows the total bacterial count of the system  $N(t)=\sum_i\sum_k\mbx_i^k=N_A+N_B$.
        The gray line denotes the dynamic of the resource pool scaled with its capacity $N_t=10^4$.
    }
    \label{fig:2pool_resource_2sp}
    \end{center}
\end{figure}
As one can see in Figure~\ref{fig:2pool_resource_2sp} even without direct interaction between species the bacterial growth was inhibited due to the limitation of the resource.
Here species $A$ was able to supress species $B$ due to the higher values of the growth constant $\alpha$.
%
\cmt{But LV for several bacteria states in literature? Didn't find!}
%
%
%
% ------------------------------------------------------------------------------------------------
\subsection{Interspecies Competition}
However, except the inhibition due to resource competition, bacteria developed different others active competition mechanisms.
One of the examples of such mechanisms is a production of the antimicrobial compounds, inhibitors or toxins~\cite{wloch-salamon_effect_2008, chao_structured_1981}.
%
%
% ................................................................................................
\subsubsection{Toxins Production}
Now extend our model where bacteria not only consume nutritions from resource pool $R$ but one of the species also actively produce the inhibiting compounds $T$. 
It is the accumulation of the bacteria waste which is responsible for the spoilage of food products. 
In order to take this into account we add an equation for the waste production:
\begin{eqnarray}
    \dot{L} &=& - \lambda_f R L + \mathcal{B}(R)G\\
    \dot{G} &=& \lambda_f R L +\alpha R G-\mathcal{B}(R)G\\
    \dot{R} &=&-\frac{\alpha}{N_t} R G\\
    \dot{T} &=& \mathcal{F}(R,G).
\end{eqnarray}
The function $\mathcal{F}$ denotes the model for the antibiotic production rate. 
We shall discuss two possible models:
\begin{eqnarray}
    \mathcal{F}_1(R,G)&=&\kappa G\\
    \mathcal{F}_2(R,G)&=&\kappa\alpha R G=-\kappa N_t\dot{R}
\end{eqnarray}
We would like to test the following model:
\begin{equation}
    \begin{cases}
        \dot{L}_A = - \lambda_A R L_A\\
        \dot{L}_B = - \lambda_B R L_B \\
        \dot{G}_A = \lambda_A R L_A + \alpha_A R G_A - \frac{\mu}{\kappa N_t} T G_A\\
        \dot{G}_B = \lambda_B R L_B + \alpha_B R G_B\\
        \dot{R} = -\frac{\alpha_A}{N_t} R G_A-\frac{\alpha_B}{N_t} R G_B\\
        \dot{T} = \mathcal{F}_{1,2} (R, G_B) \\
    \end{cases}
    \label{eq:model_2sp_toxin}
\end{equation}

As it can be seen in Figure~\ref{fig:2pool_2sp_toxin}, model $\mathcal{F}_1$ results in $I_1(t)=\kappa\int_{t_0}^tG(t')dt'$. 
As long as $G>0$ toxin is produced, no matter whether the amount of bacteria is still growing. 
Model $\mathcal{F}_2$ results in waste accumulation in steady state $I_2^s=\lim_{t\to\infty}I_1(t)=\kappa(N_t-n_0)$. 
To see this we note $\dot{T}_2=\kappa(\dot{L}+\dot{G})$, which yields: $I_2(t)=\kappa(L(t)+G(t)-n_0)$ where we used $T_2(0)=G(0)=0$, $L(0)=n_0$. 
This model assumes that inhibitor is only produced when the bacteria grow and divide.\\

@@

    \includegraphics[width=0.5\textwidth]{Figures/pool_model_1pool_inhib.pdf}

@@

    \includegraphics[width=0.67\textwidth]{Figures/pool_model_spatial.pdf}

>>>>

    \includegraphics[width=0.6\textwidth]{Figures/TPM_fig.pdf}

@@

Due to its life in rather unfavorable conditions bacteria developed different surviving mechanisms, one of which is adapting their metabolism and transitioning into dormant state.

@@

Further we show that the model can be readily extended to the systems with more than one bacterium and describe more complex relationships between species. 

@@

\label{eq:Resource}

@@

Inserting Eq.~\ref{eq:Resource} into Eqs.~\ref{eq:model_2sp_resource_comp} yields:

@@

which has the structure of a \ac{glv} expanded with different internal states for one species.
As one can see in Figure~\ref{fig:2pool_resource_2sp} even without direct interaction between species the bacterial growth was inhibited due to the limitation of the resource.
Here species $A$ has a longer lag-phase than $B$ but the exponential growth of $A$ is stronger than the one that bacteria $B$ experiences.
As a result, species $A$ was able to suppress species $B$ consuming the larger amount of nutritient and reaching larger bacterial count at the end.
    \includegraphics[width=0.6\textwidth]{Figures/pool_model_2pools_resource_competition.pdf}
        The competition of the two bacterial species due to the common limiting resource calculated using Eq.~(\ref{eq:model_2sp_resource_comp}).
        The red curve describes the growth of the bacteria $A$ ($N_A = \sum_{k} x_A^k = L_A+G_A$) with model parameters $\lambda_A=10^{-2}$, $\alpha_A=3.0$.
        The blue curve describes the growth of the bacteria $B$ ($N_B = \sum_{k} x_B^k = L_B+G_B$) with model parameters $\lambda_B=5\cdot 10^{-2}$, $\alpha_B=2.5$.
        The purple line shows the total bacterial count of the system  $N(t)=\sum_i\sum_k\mbx_i^k=N_A+N_B$.
    \label{fig:2pool_resource_2sp}
In order to get the commonly used form of the \ac{glv}, we need to simply eliminate of the bacterial states by setting $L_A=L_B=0$: 

@@

Equations (\ref{eq:LV_simple1}-\ref{eq:LV_simple2}) are basically competitive Lotka-Volterra equations that are widely used in ecological studies and show the principle of competitive excursion.
That principle claims that two species in competition for the same resource cannot coexist~\cite{wangersky_lotka-volterra_1978}. 
In previous studies it was shown that \ac{glv} equations with one bacterial state can be sucessfully applied to predict the microbial community dynamics, determine different types of the pairwise interspecies interactions~\cite{dedrick_when_2023, stein_ecological_2013, venturelli_deciphering_2018, hoffmann_power_2007}.
In food studies the \ac{glv} was applied by Mounier \etal~\cite{mounier_microbial_2008} to study interactions of six bacteria species influenced by yeast omissions in cheese.
Cauchie \etal~\cite{cauchie_modeling_2020} applied Lotka-Volterra model to study the three-species system of the bacteria majorly present in minced pork meat.\\
However, in contrast to the classical Lotka-Volterra approach, the introduced pool model allows for far more flexibility in tuning properties of the bacterial interactions within and between species. 
\cmt{But LV for several bacteria states in literature? Didn't find!}
\subsection{Interspecies Competition}
However, except the inhibition due to resource competition, bacteria developed different others active competition mechanisms.
One of the examples of such mechanisms is a production of the antimicrobial compounds, inhibitors or toxins~\cite{wloch-salamon_effect_2008, chao_structured_1981}.
On the other hand, some bacteria can produce public goods, the compounds that benefit growth of other species.
In this way, can suppress the non-preferred species by helping their competitors~\cite{hibbing_bacterial_2010}.
\subsubsection{Toxins Production}
Now extend our model where bacteria not only consume nutritions from resource pool $R$ but one of the species also actively produce the inhibiting compounds $T$. 
It is the accumulation of the bacteria waste which is responsible for the spoilage of food products. 
In order to take this into account we add an equation for the waste production:
    \dot{L} &=& - \lambda_f R L + \mathcal{B}(R)G\\
    \dot{G} &=& \lambda_f R L +\alpha R G-\mathcal{B}(R)G\\
    \dot{R} &=&-\frac{\alpha}{N_t} R G\\
    \dot{T} &=& \mathcal{F}(R,G).
The function $\mathcal{F}$ denotes the model for the antibiotic production rate. 
We shall discuss two possible models:
    \mathcal{F}_1(R,G)&=&\kappa G\\
    \mathcal{F}_2(R,G)&=&\kappa\alpha R G=-\kappa N_t\dot{R}
We would like to test the following model:
        \dot{G}_A = \lambda_A R L_A + \alpha_A R G_A - \frac{\mu}{\kappa N_t} T G_A\\
        \dot{G}_B = \lambda_B R L_B + \alpha_B R G_B\\
        \dot{R} = -\frac{\alpha_A}{N_t} R G_A-\frac{\alpha_B}{N_t} R G_B\\
        \dot{T} = \mathcal{F}_{1,2} (R, G_B) \\
    \label{eq:model_2sp_toxin}
As it can be seen in Figure~\ref{fig:2pool_2sp_toxin}, model $\mathcal{F}_1$ results in $I_1(t)=\kappa\int_{t_0}^tG(t')dt'$. 
As long as $G>0$ toxin is produced, no matter whether the amount of bacteria is still growing. 
Model $\mathcal{F}_2$ results in waste accumulation in steady state $I_2^s=\lim_{t\to\infty}I_1(t)=\kappa(N_t-n_0)$. 
To see this we note $\dot{T}_2=\kappa(\dot{L}+\dot{G})$, which yields: $I_2(t)=\kappa(L(t)+G(t)-n_0)$ where we used $T_2(0)=G(0)=0$, $L(0)=n_0$. 
This model assumes that inhibitor is only produced when the bacteria grow and divide.\\

@@

    \includegraphics[width=0.6\textwidth]{Figures/pool_model_1pool_inhib.pdf}

@@

    \includegraphics[width=0.6\textwidth]{Figures/pool_model_spatial.pdf}
